---
title: "Celltype composition with Propeller"
author: "Harvard Chan Bioinformatics Core"
date: "`r Sys.Date()`"
output:
   html_document:
      code_folding: hide
      df_print: paged
      highlights: pygments
      number_sections: true
      self_contained: true
      theme: default
      toc: true
      toc_float:
         collapsed: true
         smooth_scroll: true
params:
   project_file: ../information.R
---

```{r, message=FALSE, warning=FALSE}
# This sets up the working directory to this file so all files can be found
library(rstudioapi)
library(tidyverse)
setwd(fs::path_dir(getSourceEditorContext()$path))
# NOTE: This code will check version, this is our recommendation, it may work with other versions
stopifnot(R.version$major >= 4) # requires R4
if (compareVersion(R.version$minor, "3.1") < 0) warning("We recommend >= R4.3.1")
stopifnot(compareVersion(as.character(BiocManager::version()), "3.18") >= 0)
stopifnot(compareVersion(as.character(packageVersion("Seurat")), "5.0.0") >= 0)
```

This code is in this ![](https://img.shields.io/badge/status-draft-grey) revision.

```{r setup, cache=FALSE, message=FALSE, warning=FALSE, echo=FALSE}
# knitr::opts_chunk$set(echo = TRUE)

# Load libraries
# ... for knitting
library(knitr)
library(rmarkdown)
# ... for scRNA
library(Seurat)
library(SummarizedExperiment)
library(SingleCellExperiment)
# ... for data wrangling
library(tidyverse)
library(data.table)
library(DT)
# ... for plotting
library(DEGreport)
library(pheatmap)
library(patchwork)
library(ggprism)
library(grafify)
library(ggpubr)
# for multithreading
library(future)
# for proportional analysis
library(speckle)
library(limma)

# Set seed for reproducibility
set.seed(1454944673L)

opts_chunk[["set"]](
  audodep = TRUE,
  cache = FALSE,
  cache.lazy = FALSE,
  error = TRUE,
  echo = FALSE,
  fig.height = 5L,
  fig.retina = 2L,
  fig.width = 11,
  message = FALSE,
  tidy = TRUE,
  warning = FALSE
)

ggplot2::theme_set(theme_prism(base_size = 12))
# https://grafify-vignettes.netlify.app/colour_palettes.html
# NOTE change colors here if you wish
scale_colour_discrete <- function(...) {
  scale_colour_manual(..., values = as.vector(grafify:::graf_palettes[["kelly"]]))
}

if (future::supportsMulticore()) {
  future::plan(future::multicore)
} else {
  future::plan(future::multisession)
}
```

```{r sanitize_datatable}
sanitize_datatable <- function(df, ...) {
  # remove dashes which cause wrapping
  DT::datatable(df, ...,
    rownames = gsub("-", "_", rownames(df)),
    colnames = gsub("-", "_", colnames(df)),
    options = list(scrollX = TRUE, ...)
  )
}
```

## Load object and subset
Read in the full dataset Seurat object and subset to keep only cells from the conditions of interest.

```{r load_data}
# load seurat
seurat <- readRDS("data.rds")

# subset to conditions of interest (change to reflect your dataset)
subset_seurat <- subset(x = seurat, subset = condition %in% c("Ctrl", "Treat"))

# make sure default assay and idents are what we think they are
Idents(object = subset_seurat) <- "celltype"
DefaultAssay(subset_seurat) <- "RNA"

# Create metadata df and factor celltype
meta <- subset_seurat@meta.data
meta$celltype <- factor(meta$celltype)
```

## Count numbers of cells

```{r cell_numbers}
meta$condition_sample <- paste0(meta$condition, "_", meta$sample)
table(meta$condition_sample, meta$celltype)
```

## Run propeller

We will run `propeller` to test for differences in proportion by [condition of interest]. Briefly, the function will calculate the cell type proportions for each sample, normalize the proportions using a variance-stabilizing transformation, and fit a linear model for the effect of our variable of interest using the `limma` R package. _Note that if you have outliers, you should use the arcsin rather than the default logit transformation._ If there are only two groups, a t-test is used to calculate a p-value for each cell type that says whether the proportion of that cell type is statistically significantly different between the two groups; otherwise, the statistical test used is an ANOVA. Benjamini Hochberg false discovery rate (FDR) correction is then applied for multiple testing. (See the [`propeller` vignette, section 3](https://bioconductor.org/packages/devel/bioc/vignettes/speckle/inst/doc/speckle.html#finding-significant-differences-in-cell-type-proportions-using-propeller) for details.)

```{r run_propeller}
# working code; takes into account cell numbers
# NOTE adapt with your variables of interest

# NOTE about transformations
# see https://phipsonlab.github.io/propeller-paper-analysis/RealDataAnalysis.html#Young_vs_Old
# "While logit may be more powerful according to the simulations, it appears sensitive to outliers, and in this case we would prefer to use the arcsin square root transform."

propres <- propeller(subset_seurat,
                     sample = subset_seurat$sample,
                     clusters = subset_seurat$celltype,
                     group = subset_seurat$condition,
                     transform = "logit")

sanitize_datatable(propres)
```

## Proportions of cells table

To understand discordance with sccomp results, let's evaluate proportions and variability within group. 

```{r vst}
# NOTE about transformations
# see https://phipsonlab.github.io/propeller-paper-analysis/RealDataAnalysis.html#Young_vs_Old
# "While logit may be more powerful according to the simulations, it appears sensitive to outliers, and in this case we would prefer to use the arcsin square root transform."

props <- getTransformedProps(meta$celltype,
                             meta$condition_sample,
                             transform = "logit")

# props$Proportions
# props$TransformedProps
```

## Visualize cell proportions

```{r proportions_viz_sample, fig.height = 8, fig.width = 11}
# # by sample
# plotCellTypeProps(seurat,
#                   sample   = seurat$sample_name,         # biological replicate
#                   clusters = seurat$celltype_annotation) # cluster or celltype for every cell

# # by age
# plotCellTypeProps(seurat,
#                   sample   = seurat$age,                 # biological replicate
#                   clusters = seurat$celltype_annotation) # cluster or celltype for every cell

props_to_plot <- props$Proportions %>% as.data.frame() %>%
  # ADD YOUR METADATA: here this is done by exctracting it from the sample name
  # or you could left_join a metadata DF by "sample"
  # pull out age / sex / tissue from sample name
  separate_wider_delim(cols = sample,
                       delim = "_",
                       names = c("age", "sex", "tissue", "replicate"),
                       cols_remove = FALSE) %>%
  # clean up the proportion to use as a plot label
  mutate(Freq_label = ifelse(Freq > 0.005, round(Freq, digits = 2), NA))

# by sample -- dotplot
props_to_plot %>% ggplot(aes(x = age, y = Freq, color = sex, shape = tissue)) +
  geom_point(position = position_dodge(width = 0.2)) +
  # wrap long cluster names
  facet_grid(~ clusters, labeller = labeller(clusters = label_wrap_gen(width = 10))) +
  ggtitle("Proportions by group")

# by sample -- barplot
props_to_plot %>%
  ggplot(aes(x = sample, y = Freq, fill = clusters, label = Freq_label)) +
  geom_bar(stat = "identity") + # stacked barplot
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + # rotate x axis labels
  scale_fill_manual(values = celltype_colors) + # change colors
  geom_text(position = position_stack(vjust = 0.5), size = 2) + # add %s as labels
  facet_grid(~ age + tissue, scales = "free") +
  ggtitle("Proportions by sample")

# by age (or other factor of interest)
props_to_plot_age <- props_to_plot %>%
  # summarize frequency by age for labels
  group_by(age, clusters) %>%
  summarize(Total_Freq = sum(Freq),
            Mean_Freq  = mean(Freq)) %>%
  mutate(Total_Freq_label = ifelse(Total_Freq > 0.005, round(Total_Freq, digits = 2), NA),
         Mean_Freq_label  = ifelse(Mean_Freq  > 0.005, round(Mean_Freq,  digits = 2), NA))
plot_total <- props_to_plot_age %>%
  ggplot(aes(x = age, y = Total_Freq, fill = clusters, label = Total_Freq_label)) +
  geom_bar(stat = "identity") + # stacked barplot
  scale_fill_manual(values = celltype_colors) + # change colors
  geom_text(position = position_stack(vjust = 0.5), size = 2) + # add %s as labels
  ggtitle("Total proportion by age")
plot_average <- props_to_plot_age %>%
  ggplot(aes(x = age, y = Mean_Freq, fill = clusters, label = Mean_Freq_label)) +
  geom_bar(stat = "identity") + # stacked barplot
  scale_fill_manual(values = celltype_colors) + # change colors
  geom_text(position = position_stack(vjust = 0.5), size = 2) + # add %s as labels
  ggtitle("Mean proportion by age")
ggarrange(plot_total, plot_average, common.legend = TRUE)
```

```{r run_propeller_manually, eval=FALSE}
# get metadata
meta_propeller <- meta %>%
  group_by(condition, sample) %>%
  distinct(sample)

# set up design matrix
# NOTE: the group information is always FIRST in the design matrix specification, and there is NO intercept
# so you can add additional factors like so: designAS <- model.matrix(~ 0 + meta_propeller$age + meta_propeller$sex + meta_propeller$tissue)
designAS <- model.matrix(~ meta_propeller$condition)
colnames(designAS) <- c("Ctrl", "Treat")

# run limma manually (without contrasts)
fit <- lmFit(props$TransformedProps, designAS)
fit <- eBayes(fit, robust = TRUE)

# the p-value comes from an ANOVA testing whether the variances are the same among all the groups
# note that the other values are **log2 fold changes, not proportions as before**
topTable(fit)
```

# Resources

* [Speckle vignette](http://127.0.0.1:29356/library/speckle/doc/speckle.html#finding-significant-differences-in-cell-type-proportions-using-propeller)
* [Application to mouse PBMC scRNA](https://phipsonlab.github.io/propeller-paper-analysis/RealDataAnalysis.html#Young_vs_Old)

# Conclusion



# R session

List and version of tools used for the proportion report generation.

```{r}
sessionInfo()
```
